Revisiting the von Bertalanffy Seasonal Cessational Growth Function of Pauly et al. (1992)

### 2016-05-22 zzzOld/doc/manuscript_v0.Rmd
Moved to Word, Reorganized file strux, added mosquitofish example

Values of $C$&gt;1 (or &lt;0) in  `r eqncaps("SomersMod",display="cite")` allow seasonal decreases in mean length-at-age (`r figcaps("SomersExPlot",display="cite")`).  A decrease in mean length is unlikely for organisms whose skeletons largely preclude shrinkage [@Paulyetal1992], although a seasonal decrease in mean length-at-age is possible if size-dependent overwinter mortality occurs [@GarciaBerthouetal2012].  @Paulyetal1992 modified `r eqncaps("SomersMod",display="cite")` to include a true seasonal no-growth period where mean length was not allowed to decrease and that included a smooth transition of the modeled mean length-at-age into and out of the no-growth period.  Specifically, their modification is

@@

\noindent where $NGT$ is the "no-growth time" or the length of the no growth period (as a fraction of a year) and $t^{'}$ is found by "subtracting from the real age ($t$) the total no-growth time occurring up to age $t$" [@Paulyetal1992].  Furthermore,  @Paulyetal1992 noted that the units of $K$ changed from $year^{-1}$ in `r eqncaps("SomersMod",display="cite")` to $(1-NGT)^{-1}$ in `r eqncaps("PaulyMod",display="cite")`.  To eliminate confusion, they suggested using $K^{'}$ in `r eqncaps("PaulyMod",display="cite")`, as we do here.

@Paulyetal1992 derived  `r eqncaps("PaulyMod",display="cite")` by assuming $C$=1 (i.e., that the rate of growth is 0 at the $WP$) and replacing $2\pi$ with $\frac{2\pi}{1-NGT}$ (i.e., restricting the oscillation to the period of growth and noting that $K^{'}$ only operates during this shorter period).  Their modification may be described geometrically (though not algorithmically) in two steps.  First, the seasonal growth function in `r eqncaps("SomersMod",display="cite")` with $C$=1 is fit to the observed lengths and ages that have had the cumulative $NGT$ subtracted (i.e., using $t^{'}$).  The growth trajectory is then separated at each $WP$ and horizontal segments that are $NGT$ units long are inserted at these points.  This forms a growth trajectory that smoothly transitions into and out of the no-growth periods (`r figcaps("PaulyExPlot",display="cite")`).

@@

# The Modified Seasonal Cessation Growth Function

@@

$$
\begin{aligned}
q= & K^{'}(t^{'}-t_{0}) \\
   & +\frac{K^{'}(1-NGT)}{2\pi}sin\left(\frac{2\pi}{1-NGT}\left(t^{'}-WP-\frac{1}{2}\right)\right)  \quad \quad \quad \quad \text{`r paste0("(",eqncaps("PaulyMod2",display="num"),")")`} \\
   & -\frac{K^{'}(1-NGT)}{2\pi}sin\left(\frac{2\pi}{1-NGT}\left(t_{0}-WP-\frac{1}{2}\right)\right)
\end{aligned}  
$$

@@

As noted by @Paulyetal1992 the calculation of $t^{'}$ depends on the observed age ($t$) and the cumulative no-growth time prior to $t$.  In practice, the calculation of $t^{'}$ also depends on the position of the no-growth period within a year.  Here, the position of the no-growth period is defined by $SNG$ and $NGT$.  However, because the $SNG$ has a one-to-one relation with $WP$ and $t_{s}$, the no-growth period can also be defined relative to these parameters.

@@

# Fitting the Modified Model
Simple simulated data were initially used to demonstrate the fitting of `r eqncaps("paulyMod2",display="cite")`.  The data were simulated by randomly selecting 200 real numbers from a uniform distribution between 0 and 5 to serve as observed ages, plugging these ages into `r eqncaps("paulyMod2",display="cite")` to compute a mean length for each age, and then adding a random deviate from a normal distribution with a mean of 0 and a standard deviation of $\sigma$ to each mean length to simulate observed individual lengths.  The `nls()` function from R was then used to estimate parameter values for `r eqncaps("paulyMod2",display="cite")` from the simulated data.  The "port" algorithm in `nls()` was used as $L_{\infty}$ and $K^{'}$ were constrained to be positive and $WP$ and $NGT$ were constrained to be between 0 and 1.  One would expect the parameter estimates to be very close to the values used to create the simulated data, as `r eqncaps("paulyMod2",display="cite")` was used to both create and fit the data.

One example of a model fit to data that were simulated to have a "long" NGT, "late" WP, and low individual variability is shown in `r figcaps("PaulyExData1",display="cite")`.  In this example, the parameter estimates are quite close to the parameter values used to simulate the data (the set parameter values are within the confidence intervals for each parameter).

@@

figcaps("PaulyExData1",paste0("Example VBGF using ",eqncaps('PaulyMod2',display='cite')," with parameters shown in the upper-left corner and $\\sigma$=",sigma,".  Parameter estimates (and 95% confidence intervals) from non-linear regression are shown the lower-right corner.  Each WP is shown by the arrow and each NGT is shown by the horizontal interval bar centered on the WP arrow."))
```

@@

`r eqncaps("paulyMod2",display="cite")` was fit to simulated data sets that consisted of all combinations of "early" ($WP=0.2$) and "late" ($WP=0.8$) winter points, "short" ($NGT=0.2$) and "long" ($NGT=0.5$) no-growth periods, and "low" ($\sigma=0.3$) and "high" ($\sigma=0.5$) individual variabilities.  These limited simulations suggest that all parameters are consistently well estimated, with the possible exception of $t_{0}$.

@@

@@

@Stewartetal2013 examined the growth of 215 Bonito (*Sarda australis*) sampled from commercial landings.  Detailed methods are described in @Stewartetal2013, but note here that fork lengths (mm) were measured for each fish and ages were the decimal age calculated as the number of opaque zones observed on otolith thin sections plus the proportion of the year after the designated birthdate.  @Stewartetal2013 fit `r eqncaps("SomersMod",display="cite")` to these data but constrained $C$ to be less than 1.  Their model fit resulted in the boundary condition of $C=1$, which suggested that Bonito ceased to grow at *at least* one point.  As $C$ is at least equal to 1, this result also suggests that `r eqncaps("paulyMod2",display="cite")` should be fit to these data to determine the length of the no-growth period [@Paulyetal1992].  Thus, `r eqncaps("paulyMod2",display="cite")` is fit to these data here as an example of fitting  `r eqncaps("paulyMod2",display="cite")` to real data.

```{r echo=FALSE, results='hide'}
## Bonito data
bon <- read.csv("Bonito.csv")

################################################################################
## Purely Demonstration
## See if results match that from Stewart et al. (2013)
vbSO <- vbFuns("Somers")
lwrbnd <- c(Linf=0,K=0,t0=-Inf,C=0,ts=0)
uprbnd <- c(Linf=Inf,K=Inf,t0=Inf,C=1,ts=1)
svBon <- list(Linf=72,K=0.27,t0=0,C=0.6,ts=0.2)
fitBon1 <- nls(FL~vbSO(Age,Linf,K,t0,C,ts),data=bon,
               start=svBon,lower=lwrbnd,upper=uprbnd,
               algorithm="port",control=list(maxiter=100))
## Linf=71.9, K=0.27, t0=-1.92, C=1, and to=0.09 ... they all match (within rounding)
coef(fitBon1)
################################################################################
```

```{r echo=FALSE, results='hide', message=FALSE}
## Fit new model to Bonito data
lwrbnd <- c(Linf=0,K=0,t0=-Inf,WP=0,NGT=0)
uprbnd <- c(Linf=Inf,K=Inf,t0=Inf,WP=1,NGT=1)
svBon <- list(Linf=72,K=0.27,t0=0,WP=0.5,NGT=0.2)
fitBon2 <- nls(FL~VBSCGF(Age,Linf,K,t0,WP,NGT),data=bon,
             start=svBon,lower=lwrbnd,upper=uprbnd,
             algorithm="port",control=list(maxiter=100))
cfBon2 <- cbind(Est=coef(fitBon2),confint(fitBon2))
```

```{r echo=FALSE, results="hide"}
## A plot
par(xaxs="i",yaxs="i",mar=c(3,3,0.6,0.6),mgp=c(1.7,.5,0),tcl=-0.3,las=1)
plot(FL~Age,data=bon,pch=19,col="black",
     xlab="Age (years)",ylab="Length",xlim=c(0,4),ylim=c(0,70))
curve(VBSCGF(x,coef(fitBon2)),from=0,to=4,lwd=4,add=TRUE)
curve(vbSO(x,coef(fitBon1)),from=0,to=4,lwd=2,add=TRUE,col="gray50")
legend("bottomright",c(TeX("\\mathbf{Estimated Values}"),
                       TeX(paste0("$L_{\\infty}$"," = ",
                                  formatC(cfBon2["Linf","Est"],format="f",digits=1),
                                  " (",formatC(cfBon2["Linf","2.5%"],format="f",digits=1),
                                  ", ",formatC(cfBon2["Linf","97.5%"],format="f",digits=1),")")),
                   paste0("K' = ",formatC(cfBon2["K","Est"],format="f",digits=2),
                                  " (",formatC(cfBon2["K","2.5%"],format="f",digits=2),
                                  ", ",formatC(cfBon2["K","97.5%"],format="f",digits=2),")"),
                   TeX(paste0("$t_{0}$"," = ",
                                  formatC(cfBon2["t0","Est"],format="f",digits=2),
                                  " (",formatC(cfBon2["t0","2.5%"],format="f",digits=2),
                                  ", ",formatC(cfBon2["t0","97.5%"],format="f",digits=2),")")),
                   paste0("WP = ",formatC(cfBon2["WP","Est"],format="f",digits=2),
                                  " (",formatC(cfBon2["WP","2.5%"],format="f",digits=2),
                                  ", ",formatC(cfBon2["WP","97.5%"],format="f",digits=2),")"),
                   paste0("NGT = ",formatC(cfBon2["NGT","Est"],format="f",digits=2),
                                  " (",formatC(cfBon2["NGT","2.5%"],format="f",digits=2),
                                  ", ",formatC(cfBon2["NGT","97.5%"],format="f",digits=2),")")),
        bty="n",cex=0.8)

figcaps("PaulyBonito",paste0("Fork length at age for Australian Bonito with the best-fit of ",eqncaps('PaulyMod2',display='cite')," (black line) and ",eqncaps('SomersMod',display='cite')," (gray line) superimposed.  The parameter estimates (and 95% confidence intervals) from non-linear regression for ",eqncaps('PaulyMod2',display='cite')," are shown the lower-right corner."))
```

`r figcaps("PaulyBonito")`

@@

* General -- fits perfectly simulated data well.
* Assumptions -- WP same time each year and age, NGT same length each year and age
* Model-Fitting -- bounding parameters, more difficult to fit five parameters
* Practical Difference -- not large unless C>>1 and NGT>>0

>>>>

Values of $C$&gt;1 (or &lt;0) in  `r eqncaps("SomersMod",display="cite")` allow seasonal decreases in mean length-at-age (`r figcaps("SomersExPlot",display="cite")`).  A decrease in mean length is unlikely for organisms whose skeletons largely preclude shrinkage [@Paulyetal1992], although a seasonal decrease in mean length-at-age is possible if size-dependent overwinter mortality occurs [@GarciaBerthouetal2012].  To address this unrealistic decrease in mean length, @Paulyetal1992 modified `r eqncaps("SomersMod",display="cite")` to include a true seasonal no-growth period that included a smooth transition of the modeled mean length-at-age into and out of the no-growth period.  Specifically, their modification is

@@

\noindent where $NGT$ is the "no-growth time" or the length of the no growth period (as a fraction of a year) and $t^{'}$ is found by "subtracting from the real age ($t$) the total no-growth time occurring up to age $t$."  Furthermore,  @Paulyetal1992 noted that the units of $K$ changed from $year^{-1}$ in `r eqncaps("SomersMod",display="cite")` to $(1-NGT)^{-1}$ in `r eqncaps("PaulyMod",display="cite")`.  To eliminate confusion, he suggested using $K^{'}$ in `r eqncaps("PaulyMod",display="cite")`, as we do here.

@Paulyetal1992 derived  `r eqncaps("PaulyMod",display="cite")` by assuming $C$=1 (i.e., that the rate of growth is 0 at the $WP$) and replacing $2\pi$ with $\frac{2\pi}{1-NGT}$ (i.e., restricting the oscillation to the period of growth and noting that $K^{'}$ only operates in this shorter period).  Their modification may be described geometrically (though not algorithmically) in two steps.  First, the seasonal growth function in `r eqncaps("SomersMod",display="cite")` with $C$=1 is fit to the observed lengths and ages that have had the cumulative $NGT$ subtracted (i.e., using $t^{'}$).  The growth trajectory is then separated at each $WP$ and horizontal segments that are $NGT$ units long are inserted at these points.  This forms a growth trajectory that smoothly transitions into and out of the no-growth periods (`r figcaps("PaulyExPlot",display="cite")`).

@@

# The Modified Model

@@

$$
\begin{aligned}
q= & K^{'}(t^{'}-t_{0}) \\
   & +\frac{K^{'}(1-NGT)}{2\pi}sin\left(\frac{2\pi}{1-NGT}\left(t^{'}-WP-\frac{1}{2}\right)\right)-\frac{K^{'}(1-NGT)}{2\pi}sin\left(\frac{2\pi}{1-NGT}\left(t_{0}-WP-\frac{1}{2}\right)\right)  \quad \quad \quad \quad \text{`r paste0("(",eqncaps("PaulyMod2",display="num"),")")`}
\end{aligned}  
$$

@@

As noted by @Paulyetal1992 the calculation of $t^{'}$ depends on the observed age ($t$) and the cumulative no-growth time prior to $t$.  In practice, the calculation of $t^{'}$ also depends on the position of the no-growth period within a year.  Here, the position of the no-growth period is defined by $SNG$.  However, because the $SNG$ has a one-to-one relation with $WP$ and $t_{s}$ it can also be defined relative to these parameters.

@@

# Fitting the Modified Model
Simple simulated data were used to demonstrate the fitting of `r eqncaps("paulyMod2",display="cite")`.  The data were simulated by randomly selecting 200 real numbers from a uniform distribution between 0 and 5 to serve as observed ages, plugging these ages into `r eqncaps("paulyMod2",display="cite")` to compute a mean length for each age, and then adding a random deviate from a normal distribution with a mean of 0 and a standard deviation of $\sigma$ to each mean length to simulate observed individual lengths.  The `nls()` function from R was then used to estimate parameter values for `r eqncaps("paulyMod2",display="cite")` from the simulated data.  The "port" algorithm in `nls()` was used as $L_{\infty}$ and $K^{'}$ were constrained to be positive and $WP$ and $NGT$ were constrained to be between 0 and 1.  One would expect the parameter estimates to be very close to the values used to create the simulated data as `r eqncaps("paulyMod2",display="cite")` was used to both create and fit the data.

One example of a model fit to data that were simulated to have a "long" NGT, "late" WP, and low individual variability is shown in `r figcaps("PaulyExData1",display="cite")`.  In this example, the parameter estimates are quite close to the parameter values used to simulate the data (the set parameter values are within the confidence interval for each parameter).

@@

figcaps("PaulyExData1",paste0("Example VBGF using ",eqncaps('PaulyMod2',display='cite')," with parameters shown in the upper-left corner and $\\sigma$=",sigma,".  Paramter estimates (and 95% confidence intervals) from non-linear regression are shown the lower-right corner.  Each WP is shown by the arrow and each NGT is shown by the horizontal interval bar centered on the WP arrow."))
```

@@

@@

```{r echo=FALSE, results='hide',eval=FALSE}
## Set some common parameters
Linf <- 30; K <- 0.3; t0 <- -0.1
## Set bounds for model-fitting
lwrbnd <- c(Linf=0,K=0,t0=-Inf,WP=0,NGT=0)
uprbnd <- c(Linf=Inf,K=Inf,t0=Inf,WP=1,NGT=1)
```

```{r echo=FALSE, results="hide", message=FALSE,eval=FALSE}
## Setup
par(xaxs="i",yaxs="i",mar=c(3,3,0.6,0.6),mgp=c(1.7,.5,0),tcl=-0.3,las=1)
## Set some specific growth parameters
# fairly typical north temperate situation
WP <- 0.8; NGT <- 0.5
gper <- c(WP+NGT/2-1,WP-NGT/2)*365
## Actually generate the data (with very little variability)
d1 <- vbDataGen(200,Linf=Linf,K=K,t0=t0,maxAge=5,SE=0.001,SD=0.001,paramCV=0.001,
                growth.per=gper,sample.per=c(1,365),lendigs=1)
## Try fitting the VBSCGF model
sv <- list(Linf=32,K=0.32,t0=0,WP=0.9,NGT=0.3)
fitP1 <- nls(lenCap~VBSCGF(ageFracY,Linf,K,t0,WP,NGT),data=d1,
             start=sv,lower=lwrbnd,upper=uprbnd,
             algorithm="port",control=list(maxiter=100))
( cfs <- coef(fitP1) )
## A plot
plot(lenCap~ageFracY,data=d1,pch=19,col=col2rgbt("black",1/3),
     xlab="Age (years)",ylab="Length",xlim=c(1,6),ylim=c(7,27))
curve(VBSCGF(x,coef(fitP1)),from=1,to=6,lwd=2,add=TRUE)
WPs <- 1:5+cfs[["WP"]]
LatWPs <- VBSCGF(WPs,Linf=cfs)
arrows(WPs,LatWPs+2,WPs,LatWPs+0.2,lwd=2,length=0.1)
SNGTs <- WPs-cfs[["NGT"]]/2
ENGTs <- WPs+cfs[["NGT"]]/2
arrows(SNGTs,LatWPs-0.7,ENGTs,LatWPs-0.7,lwd=2,length=0.05,angle=90,code=3)

tmp <- c("Linf=","K=","t0=","WP=","NGT=")
legend("topleft",paste0(c("Set Values",tmp),c("",Linf,K,t0,WP,NGT)),
       bty="n",inset=-0.02,cex=0.8)
cfs <- round(cbind(Est=coef(fitP1),confint(fitP1)),3)
legend("bottomright",c("Estimated Values",paste0(tmp,cfs[,1]," (",cfs[,2],",",cfs[,3],")")),bty="n",cex=0.8)
```

```{r echo=FALSE, results="hide", message=FALSE,eval=FALSE}
## Setup
par(xaxs="i",yaxs="i",mar=c(3,3,0.6,0.6),mgp=c(1.7,.5,0),tcl=-0.3,las=1)
## Set some specific growth parameters
# fairly typical north temperate situation
WP <- 0.8; NGT <- 0.5
gper <- c(WP+NGT/2-1,WP-NGT/2)*365
## Actually generate the data (with more variability)
d1 <- vbDataGen(200,Linf=Linf,K=K,t0=t0,maxAge=5,SE=0.2,SD=0.5,paramCV=0.01,
                growth.per=gper,sample.per=c(1,365),lendigs=1)
## Try fitting the VBSCGF model
sv <- list(Linf=32,K=0.32,t0=0,WP=0.9,NGT=0.3)
fitP1 <- nls(lenCap~VBSCGF(ageFracY,Linf,K,t0,WP,NGT),data=d1,
             start=sv,lower=lwrbnd,upper=uprbnd,
             algorithm="port",control=list(maxiter=100))
( cfs <- coef(fitP1) )
## A plot
plot(lenCap~ageFracY,data=d1,pch=19,col=col2rgbt("black",1/3),
     xlab="Age (years)",ylab="Length",xlim=c(1,6),ylim=c(7,27))
curve(VBSCGF(x,coef(fitP1)),from=1,to=6,lwd=2,add=TRUE)
WPs <- 1:5+cfs[["WP"]]
LatWPs <- VBSCGF(WPs,Linf=cfs)
arrows(WPs,LatWPs+2,WPs,LatWPs+0.2,lwd=2,length=0.1)
SNGTs <- WPs-cfs[["NGT"]]/2
ENGTs <- WPs+cfs[["NGT"]]/2
arrows(SNGTs,LatWPs-0.7,ENGTs,LatWPs-0.7,lwd=2,length=0.05,angle=90,code=3)

tmp <- c("Linf=","K=","t0=","WP=","NGT=")
legend("topleft",paste0(c("Set Values",tmp),c("",Linf,K,t0,WP,NGT)),
       bty="n",inset=-0.02,cex=0.8)
cfs <- round(cbind(Est=coef(fitP1),confint(fitP1)),3)
legend("bottomright",c("Estimated Values",paste0(tmp,cfs[,1]," (",cfs[,2],",",cfs[,3],")")),bty="n",cex=0.8)
```

```{r echo=FALSE, results="hide", message=FALSE,eval=FALSE}
## Setup
par(xaxs="i",yaxs="i",mar=c(3,3,0.6,0.6),mgp=c(1.7,.5,0),tcl=-0.3,las=1)
## Set some specific growth parameters
# fairly typical north temperate situation
WP <- 0.8; NGT <- 0.5
gper <- c(WP+NGT/2-1,WP-NGT/2)*365
## Actually generate the data (with more variability)
d1 <- vbDataGen(200,Linf=Linf,K=K,t0=t0,maxAge=5,SE=0.2,SD=0.5,paramCV=0.01,
                growth.per=gper,sample.per=c(1,365),lendigs=1)
## Try fitting the VBSCGF model
sv <- list(Linf=32,K=0.32,t0=0,WP=0.9,NGT=0.3)
fitP1 <- nls(lenCap~VBSCGF(ageFracY-0.3,Linf,K,t0,WP,NGT),data=d1,
             start=sv,lower=lwrbnd,upper=uprbnd,
             algorithm="port",control=list(maxiter=100))
( cfs <- coef(fitP1) )
## A plot
plot(lenCap~I(ageFracY-0.3),data=d1,pch=19,col=col2rgbt("black",1/3),
     xlab="Age (years)",ylab="Length",xlim=c(1,6),ylim=c(7,27))
curve(VBSCGF(x,coef(fitP1)),from=1,to=6,lwd=2,add=TRUE)
WPs <- 1:5+cfs[["WP"]]
LatWPs <- VBSCGF(WPs,Linf=cfs)
arrows(WPs,LatWPs+2,WPs,LatWPs+0.2,lwd=2,length=0.1)
SNGTs <- WPs-cfs[["NGT"]]/2
ENGTs <- WPs+cfs[["NGT"]]/2
arrows(SNGTs,LatWPs-0.7,ENGTs,LatWPs-0.7,lwd=2,length=0.05,angle=90,code=3)

tmp <- c("Linf=","K=","t0=","WP=","NGT=")
legend("topleft",paste0(c("Set Values",tmp),c("",Linf,K,t0,WP-0.3,NGT)),
       bty="n",inset=-0.02,cex=0.8)
cfs <- round(cbind(Est=coef(fitP1),confint(fitP1)),3)
legend("bottomright",c("Estimated Values",paste0(tmp,cfs[,1]," (",cfs[,2],",",cfs[,3],")")),bty="n",cex=0.8)
```

@@

@@
